---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# semEff

<!-- badges: start -->

[![](https://www.r-pkg.org/badges/version/semEff?color=blue)](https://CRAN.R-project.org/package=semEff) [![Downloads](https://cranlogs.r-pkg.org/badges/semEff)](https://cran.rstudio.com/package=semEff) [![GPLv3 license](https://img.shields.io/badge/License-GPL3-green.svg)](https://www.gnu.org/licenses/gpl-3.0.en.html) [![R-CMD-check](https://github.com/murphymv/semEff/workflows/R-CMD-check/badge.svg)](https://github.com/murphymv/semEff/actions) [![Donate](https://img.shields.io/badge/PayPal-Donate%20to%20Author-yellow.svg)](https://paypal.me/murphymv1)

<!-- badges: end -->

`semEff` provides functionality to automatically calculate direct, indirect, and total effects for 'piecewise' structural equation models, comprising lists of fitted models representing structured equations (Shipley 2000, 2009; Lefcheck 2016). Confidence intervals are provided via bootstrapping.

Currently supported model classes are `"lm"`, `"glm"`, `"lmerMod"`, `"glmerMod"`, `"lmerModLmerTest"`, `"gls"`, and `"betareg"`.

## Installation

You can install the released version of `semEff` from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("semEff")
```

And the development version from [GitHub](https://github.com/) with:

``` r
devtools::install_github("murphymv/semEff@dev")
```

## Example

```{r example}
# install.packages(c("semEff", "ggplot2"))
library(semEff)
library(ggplot2)

# Simulated data from Shipley (2009) on tree growth and survival (see ?Shipley)
head(shipley)

# Hypothesised SEM: latitude -> degree days to bud burst -> date of burst -> growth -> survival
lapply(shipley.sem, formula)

# Bootstrap model effects (10,000 reps... can take a while)
# system.time(
#   shipley.sem.boot <- bootEff(shipley.sem, R = 10000, seed = 53908, ran.eff = "site")
# )

# Calculate SEM effects (use saved bootstrapped SEM)
(eff <- semEff(shipley.sem.boot))

# Effects and CIs for response "Growth"
summary(eff, "Growth")

# Extract total effects for Growth
tot <- totEff(eff, "Growth")
tot.b <- totEff(eff, "Growth", type = "boot")

# Predict effects for "Date" (direct) and "DD" (indirect) on Growth
mod <- shipley.sem$Growth
dat <- na.omit(shipley)
fit <- sapply(c("Date", "DD"), function(i) {
  x <- seq(min(dat[i]), max(dat[i]), length = 100)
  x <- data.frame(x); names(x) <- i
  f <- predEff(mod, newdata = x, effects = tot[i], eff.boot = tot.b)
  c(x, f)
}, simplify = FALSE)

# Function to plot predictions
plotFit <- function(x, y, fit, x.lab = NULL, y.lab = NULL) {
  x2 <- fit[[1]]; f <- fit$fit; ci.l <- fit$ci.lower; ci.u <- fit$ci.upper
  ggplot () + 
    geom_point(aes(x, y)) +
    geom_ribbon(aes(x2, ymin = ci.l, ymax = ci.u, alpha = "0.15"), fill = "blue") +
    geom_line(aes(x2, f), color = "blue", size = 1) +
    xlab(x.lab) + 
    ylab(y.lab) +
    theme_bw() + 
    theme(legend.position = "none")
}

# Direct effects of Date
plotFit(x = dat$Date, y = dat$Growth, fit = fit$Date, x.lab = "Date of Bud Burst", y.lab = "Stem Growth")

# Indirect effects of DD (operating via Date)
plotFit(x = dat$DD, y = dat$Growth, fit = fit$DD, x.lab = "Degree Days to Bud Burst", y.lab = "Stem Growth")

# Huge amount of scatter around each fit as random effects explain most variation in stem growth! 
# Compare conditional vs. marginal R-squared:
r2 <- c(R2_cond = R2(mod)[[1]], R2_marg = R2(mod, re.form = NA)[[1]])
round(r2, 3)
```

## References

Lefcheck, J. S. (2016). piecewiseSEM: Piecewise structural equation modelling in R for ecology, evolution, and systematics. *Methods in Ecology and Evolution*, **7**(5), 573–579. <https://doi.org/f8s8rb>

Shipley, B. (2000). A New Inferential Test for Path Models Based on Directed Acyclic Graphs. *Structural Equation Modeling: A Multidisciplinary Journal*, **7**(2), 206–218. <https://doi.org/cqm32d>

Shipley, B. (2009). Confirmatory path analysis in a generalized multilevel context. *Ecology*, **90**(2), 363–368. <https://doi.org/bqd43d>
