---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# semEff

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/murphymv/semEff.svg?branch=master)](https://travis-ci.org/murphymv/semEff)
<!-- badges: end -->

semEff provides functionality to automatically calculate direct, indirect, and
total effects from 'piecewise' structural equation models, comprising lists of
fitted models representing structured equations (Shipley 2000, 2009; Lefcheck
2016). Confidence intervals are provided via bootstrapping.

Currently supported model classes are "lm", "glm", "lmerMod", "glmerMod", "gls",
and "lm.gls".

## Installation

You can install the released version of semEff from
[CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("semEff")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("murphymv/semEff@develop")
```
## Example

```{r example}
# install.packages(c("semEff", "ggplot2"))
library(semEff)
library(ggplot2)

## Simulated data from Shipley (2009) on tree growth and survival (see ?Shipley)
head(Shipley)

## Hypothesised SEM:
## latitude -> degree days to bud burst -> date of burst -> growth -> survival
lapply(Shipley.SEM, formula)

# ## Bootstrap model effects (takes a while...)
# system.time(
#   Shipley.SEM.Boot <- bootEff(Shipley.SEM, ran.eff = "site", seed = 53908)
# )

## Calculate SEM effects and CI's (use bootstrapped SEM)
eff <- suppressWarnings(semEff(Shipley.SEM.Boot))

## Summary of effects for response "Growth"
eff$Summary$Growth

## Extract total effects for Growth
tot <- totEff(eff)[["Growth"]]
tot.b <- totEff(eff, type = "boot")[["Growth"]]

## Predict effects for "Date" (direct) and "DD" (indirect) on Growth
mod <- Shipley.SEM$Growth
dat <- na.omit(Shipley)
fit <- sapply(c("Date", "DD"), function(i) {
  x <- data.frame(seq(min(dat[i]), max(dat[i]), length = 100)); names(x) <- i
  c(x, predEff(mod, newdata = x, effects = tot[i], eff.boot = tot.b))
}, simplify = FALSE)

## Function to plot predictions
plotFit <- function(x, y, fit, x.lab = NULL, y.lab = NULL) {
  x2 <- fit[[1]]; f <- fit[[2]]; ci.l <- fit[[3]]; ci.u <- fit[[4]]
  ggplot () + 
    geom_point(aes(x, y)) +
    geom_ribbon(aes(x2, ymin = ci.l, ymax = ci.u), fill = "blue", alpha = "0.15") +
    geom_line(aes(x2, f), color = "blue", size = 1) +
    xlab(x.lab) + ylab(y.lab) +
    theme_bw() + theme(legend.position = "none")
}

## Direct effects of Date
plotFit(x = dat$Date, y = dat$Growth, fit = fit$Date, 
        x.lab = "Date of Bud Burst", y.lab = "Stem Growth")

## Indirect effects of DD (operating via Date)
plotFit(x = dat$DD, y = dat$Growth, fit = fit$DD, 
        x.lab = "Degree Days to Bud Burst", y.lab = "Stem Growth")

## Huge amount of scatter around each fit as random effects explain most
## variation in this model! Compare conditional vs. marginal R-squared:
round(c(R2c = R2(mod)[[1]], R2m = R2(mod, re.form = NA)[[1]]), 3)
```

## References

* Lefcheck, J. S. (2016). piecewiseSEM: Piecewise structural equation modelling
in r for ecology, evolution, and systematics. Methods in Ecology and Evolution,
7(5), 573–579. https://doi.org/f8s8rb
* Shipley, B. (2000). A New Inferential Test for Path Models Based on Directed
Acyclic Graphs. Structural Equation Modeling: A Multidisciplinary Journal, 7(2),
206–218. https://doi.org/cqm32d
* Shipley, B. (2009). Confirmatory path analysis in a generalized multilevel
context. Ecology, 90(2), 363–368. https://doi.org/bqd43d
