---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# semEff

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/murphymv/semEff.svg?branch=master)](https://travis-ci.org/murphymv/semEff)
<!-- badges: end -->

semEff provides functionality to automatically calculate direct, indirect, and total effects from 'piecewise' Structural Equation Models, comprising lists of fitted models representing structured equations. Confidence intervals are provided via bootstrapping. 

Currently supported model classes are "lm", "glm", and "merMod".

## Installation

You can install the released version of semEff from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("semEff")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("murphymv/semEff")
```
## Example

```{r example}
# install.packages(c("semEff", "ggplot2"))
library(semEff)
library(ggplot2)

## Simulated data from Shipley (2009) (?Shipley)
head(Shipley)

## Hypothesised SEM
lapply(Shipley.SEM, formula)

# ## Bootstrap model effects (will take a while...)
# system.time(
#   Shipley.SEM.Boot <- bootEff(Shipley.SEM, ran.eff = "site", seed = 53908)
# )

## Calculate SEM effects and CI's (use saved bootstrapped SEM)
eff <- suppressWarnings(semEff(Shipley.SEM.Boot))

## Summary of effects for response "Growth"
eff$Summary$Growth

## Extract total effects for "Growth"
tot <- totEff(eff, type = "orig")[["Growth"]]
tot.b <- totEff(eff, type = "boot")[["Growth"]]

## Predict effects for "Date" (direct) and "DD" (indirect) on "Growth"
mod <- Shipley.SEM$Growth
dat <- na.omit(Shipley)
fit <- sapply(c("Date", "DD"), function(i) {
  x <- data.frame(seq(min(dat[i]), max(dat[i]), length = 100))
  names(x) <- i
  c(x, predEff(mod, newdata = x, effects = tot[i], eff.boot = tot.b))
}, simplify = FALSE)

## Function to plot predictions
plotFit <- function(x, y, fit, x.lab = NULL, y.lab = NULL) {
  x2 <- fit[[1]]; f <- fit[[2]]; ci.l <- fit[[3]]; ci.u <- fit[[4]]
  ggplot () +
  geom_point(aes(x, y)) +
  geom_ribbon(aes(x2, ymin = ci.l, ymax = ci.u), fill = "blue", alpha = "0.15") +
  geom_line(aes(x2, f), color = "blue", size = 1) +
  xlab(x.lab) + ylab(y.lab) +
  theme_bw() + theme(legend.position = "none")
}

## Direct effects of "Date"
plotFit(x = dat$Date, y = dat$Growth, fit = fit$Date, 
        x.lab = "Date of Bud Burst", y.lab = "Stem Growth")

## Indirect effects of "DD" (operating via "Date")
plotFit(x = dat$DD, y = dat$Growth, fit = fit$DD, 
        x.lab = "Degree Days to Bud Burst", y.lab = "Stem Growth")

## Huge amount of scatter around each fit as random effects explain most
## variation in this model! Compare conditional vs. marginal R-squared:
round(c(R2c = R2(mod)[[1]], R2m = R2(mod, re.form = NA)[[1]]), 3)
```
